State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-08-27 analysis/paper/paper.Rmd
added abstract.
deleted a few comments in the text.

Slight revision to the methods description of which fish had genetic samples taken.

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
- AckermanLibrary.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  This is the abstract
keywords: |
  escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also show that observed nighttime passage and fallback/re-ascension rates are typically unequal. Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical]. To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip. Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension. Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.

<!-- RK comment: the following sentence is better left in the discussion and can be removed from this paragraph. -- Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure. -->

@@

All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length. For adipose-intact (unclipped) adults, which includes wild and hatchery-no clip individuals, some portion or all of fish trapped had scale and genetic tissue samples taken. Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012a] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019). Starting with adults spawning in 2013, every unclipped Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin. Previous to that, only fish determined to be wild in origin at the trap were sampled for scale and genetics. @Camacho2017 provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks and tags, including genetic parentage-based tags [@Steele2013; @Steele2019]. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2017].

@@

<!--The true trapping rate was estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per. comm.*). Therefore, we use the mark-recapture estimate of trap rate. -->

<!-- Don't you think it's important to explain why we need to estimate the trap rate? Because otherwise after reading the description of the adult fish trap data, I would assume the trap rate is known -->

@@

We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Coefficients of variation ranged from 2.7-24.8% for wild fish, 2.2-10.5% for hatchery fish, 3.6-11% for hatchery no-clip fish and 2.1-10.4% for total unique fish past Lower Granite Dam.

<!-- 
All of the high values were from 2019, and the ranges drop quite a bit if we exclude that year. Max is more like 6%. Worth trying to mention that?

MA: Maybe worth including in the Discussion?
-->

@@

p1 = ggplot(data.frame(x = c(-5, 8)),
            aes(x)) +
  stat_function(fun = dnorm,
                args = list(mean = 4, sd = 0.4),
                aes(color = 'Daytime Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                geom = 'blank') +
  scale_color_brewer(palette = 'Set1') +
  geom_vline(xintercept = 4,
             linetype = 2) +
  coord_cartesian(xlim = my_xlim,
                  ylim = my_ylim) +
  theme_classic() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = 'bottom') +
  labs(x = 'Escapement',
       y = 'Density',
       color = 'Fish Type',
       title = 'Daytime Fish')

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
- AckermanLibrary.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of the number of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successfully management of fisheries in that area. Here we describe a state-space model that estimates such escapement past a dam by using window counts, PIT tag observations and data from an adult fish trap, accounting for issues such as night-time passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested this approach using a simulation framework that mimicked several levels of observation error, differences between night-time passage and re-ascension rates and the possibility of the adult trap being closed for some period of time, and found it to produce unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite dam to produces estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also show that observed nighttime passage and fallback/re-ascension rates are typically unequal. Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical]. To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip. Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension. Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. We then applied this model to Chinook Salmon and steelhead returns at Lower Granite dam for spawn years 2010 - 2019. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.

@@

All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length. For adipose-intact (unclipped) adults, which includes wild and hatchery-no clip individuals, some portion or all of fish trapped had scale and genetic tissue samples taken. Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012a] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019). Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics. Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin. @Camacho2017 provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks and tags, including genetic parentage-based tags [@Steele2013; @Steele2019]. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2017].

@@

@@

We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Coefficients of variation ranged from 2.7-24.8% for wild fish, 2.2-10.5% for hatchery fish, 3.6-11% for hatchery no-clip fish and 2.1-10.4% for total unique fish past Lower Granite Dam. Excluding 2019, which had the lowest escapement estimates for both species, the maximum CVs were 6.6% for wild fish, 5.6% for clipped hatchery fish, 8% for unclipped hatchery fish and 5.2% for total unque fish.

@@

p1 = ggplot(data.frame(x = c(-5, 8)),
            aes(x)) +
  stat_function(fun = dnorm,
                args = list(mean = 4, sd = 0.4),
                aes(color = 'Daytime Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                geom = 'blank') +
  scale_color_brewer(palette = 'Set1') +
  # scale_color_grey(start = 0.9,
  #                  end = 0.3) +
  geom_vline(xintercept = 4,
             linetype = 2) +
  coord_cartesian(xlim = my_xlim,
                  ylim = my_ylim) +
  theme_classic() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = 'bottom') +
  labs(x = 'Escapement',
       y = 'Density',
       color = 'Fish Type',
       title = 'Daytime Fish')
